From physical principles to ???

### 2016-04-15 2015CompleteDerivation/completeDerivation.tex
Starting to tackle classical phase space

In this context, we call the evolution of the system \emph{deterministic} if the state at a given time uniquely identifies states at future times, and \emph{reversible} if it uniquely identifies states at past times. While this is a common enough definition, we need to be clear how this applies to the unstated part. That is, the evolution of the unstated part under the same process is \emph{always} non-deterministic (and non-reversible) in the sense that the state of the system does not determine its evolution. For example, suppose we define the state as position and momentum of the center of mass of a cannonball. Suppose that the evolution is deterministic on that state. What does it tell us about the cannonball temperature, or about the motion of each of its atoms? Nothing. In this sense, the unstated part is non-deterministic and non-reversible. Could we extend the state and the laws of evolution to account for temperature? Yes, but that would be a different process defined on a different state. Does it mean the unstated part is always evolving chaotically? Not at all. The temperature may remain constant throughout the motion of the cannonball. Yet we wouldn't know, since we are not studying it: the motion is deterministic and/or reversible only as far as the state is concerned. With this in mind, we introduce the following:

@@

Suppose we study the motion of a cannonball; its state under gravitational and inertial forces will be properly described by the position and momentum of the center of mass. While light and air molecules may scatter off its surface unpredictably, its trajectory is not greatly affected as it is a massive rigid body. Suppose we study the motion of a small particle, small enough that the random scattering does influence the trajectory and it undergoes Brownian motion: its state will be a probability distribution for position and momentum of the center of mass. Gravitational and inertial forces have not changed, yet the set of states has. In other words, the set of states must also be closed under non-deterministic evolution. If the Brownian motion is not negligible, we do not end in a well defined position/momentum pair, even if we start from one.

@@

As we have hinted, sometimes the state is identified by a distribution (either statistical or actual). Even in this case, the state can be deterministic and reversible. That is, given the distribution at one time we can determine the distribution at future times. The shape and the parameters of the distribution can be deterministic, even if the trajectories of the parts are not as they fall within the unstated part. Note that we \emph{cannot} assume trajectories and states are always defined for the unstated part, as this includes also the unknown unknowns. We will return to this aspect when discussing quantum systems.

@@

We now proceed to characterize states and physically distinguishability in more precise terms so that we can capture their description mathematically. What we'll see is that the outcomes of all physical processes that can be used to gain information about the system, that can be used to perform a measurement, induce a topology on the set of states that is Hausdorff and second countable. That is, physical distinguishability is mathematically captured by topological distinguishability. Deterministic and reversible evolution will then preserve the topology, and they will be mathematically captured by self-homeomorphisms.

@@

We can also conceptually combine two different processes into a single one. That is, having a way to measure quantity $x$ and a way to measure quantity $y$ gives us a way to measure the combination $(x,y)$. For ensembles, if we can measure the marginal distribution $\rho_x(x)$ and the marginal distribution $\rho_y(x)$, we know the joint distribution $\rho(x,y)$ has to be compatible with both. Note, though, that this does not provide a way to fully measure $\rho(x,y)$ as we know nothing about the correlation between the two variables.\footnote{The quantum case is similar: we measure marginal distributions and rule out states that are incompatible with them.} Formally, the states compatible with the outcomes of the combined process will be the intersections of the states compatible with each pair of outcomes of the original processes.

@@

\begin{justification}
We claim $\mathcal{S}$ is a set. Each state is well defined as it is physically distinguishable. The collection of all possible states form a set.

We claim $\mathcal{S}$ has a topology $\mathsf{T}$. Consider the set of all possible physical outcomes associated with all physical processes. Each possible outcome is associated with a set of states that are compatible with that outcome. Let $\mathsf{T}$ be the set of all sets associated with all physical outcomes. $\varnothing \in \mathsf{T}$ and is associated with impossible outcomes. $\mathcal{S} \in \mathsf{T}$ and is associated with unavoidable outcomes. Let $V_1, V_2 \in \mathsf{T}$. Then, by definition, there exist a process $P_1$ that admits $V_1$ as an outcome and a process $P_2$ that admits $V_2$ as an outcome. Consider the process $P$ that combines the outcomes of $P_1$ and $P_2$. This always exist physically as we can prepare the same state multiple times and let it interact with each process separately. $P$ will have a possible outcome $V$ corresponding to the case where $P_1$ gave outcome $V_1$ and $P_2$ gave outcome $V_2$. The states compatible with $V$ must be in both $V_1$ and $V_2$, that is $V = V_1 \cap V_2$. Therefore $\mathsf{T}$ is closed under intersection. Let $V_1, V_2 \in \mathsf{T}$. If they are physically distinguishable, then there exists a physical process $P$ that admits both as outcomes. Given $P$, we can always construct the process $P_0$ that combines $V_1, V_2$ into a single outcome $V$ by "forgetting" which of the two was given. The states compatible with $V$ must be in either $V_1$ or $V_2$, that is $V = V_1 \cup V_2$. Therefore $\mathsf{T}$ is closed under union. $\mathsf{T}$ is a topology by definition.

@@

We claim that $\mathcal{S}$ is second countable. Consider the set of outcomes of a process. As each is identified by a finite number of digits, the set of outcomes is countable. This remains true when combining processes: the number of processes we can combine is finite as is the number of times we can prepare the same state and the intersections of countably many sets is still countable. Consider a basis for the topology. Suppose it is of cardinality greater than countable. Then no physical process could ever distinguish between the elements of the basis. The basis, and therefore the outcomes, would not be physically distinguishable. The basis for the topology cannot be of cardinality greater than countable. By contradiction, $\mathcal{S}$ must be second countable.

@@

Note that the arguments that lead to the topological space had nothing to do with states per se, just that they are physically distinguishable. States, though, are not the only objects with that property. In fact: any element of a set of physical objects (e.g. forces, physical properties such as mass or charge, time) needs to be physically distinguishable to be well defined. We can generalize the above justification: any set of physically distinguishable elements is a topological space.

@@

Note that the isomorphism is to a subspace of continuous functions as we may have to exclude, for example, those that are not bounded or those that do not vanish at infinity. These may, in some cases, not represent physically meaningful system.

@@

Math: need linear functional from $\rho$ to total amount of material. For Riesz representation theorem, there can be only one. Density must be invariant under coordinate transformation. Density live on $T^*\mathbbm{Q}$. Must be able to define marginal distributions on submanifold. Metric is two form. Metric must be form invariant under coord transformation. Metric is canonical two form.

Physics, 1D, show that densities do not work on Q. Show they work on T*Q. Define degree of freedom. Independent degree of freedom are orthogonal. Area in each d.o.f. is number of possibilities. Total number of states is product of number of possibilities for independent d.o.f. Independent d.o.f. are orthogonal. Leads to omega metric. Density allows to count particle states. Example: two uniform distribution for the same amount of particles.

@@

\begin{defn}\label{classical_phase_space}
The state space $\mathcal{S}$ for the particles of a homogeneous classical  system is a cotangent bundle $T^*\mathbbm{Q}$ with symplectic metric $\omega$.
\end{defn}

@@

>>>>

In this context, we call the evolution of the system \emph{deterministic} if the state at a given time uniquely identifies states at future times, and \emph{reversible} if it uniquely identifies states at past times. While this is a common enough definition, we need to be clear how this applies to the unstated part. That is, the evolution of the unstated part under the same process is \emph{always} non-deterministic (and non-reversible) in the sense that the state of the system does not determine its evolution. For example, suppose we define the state as position and momentum of the center of mass of a cannonball. Suppose that the evolution is deterministic on that state. What does it tell us about the cannonball temperature, or about the motion of each of its atoms? Nothing. In this sense, the unstated part is non-deterministic and non-reversible. Could we extend the state and the laws of evolution to account for temperature? Yes, but that would be a different evolution defined on a different state. Does it mean the unstated part is always evolving chaotically? Not at all. The temperature may remain constant throughout the motion of the cannonball. Yet we wouldn't know, since we are not studying it: the evolution is deterministic and/or reversible only as far as the state is concerned. With this in mind, we introduce the following:

@@

Suppose we study the motion of a cannonball; its state under gravitational and inertial forces will be properly described by the position and momentum of the center of mass. While light and air molecules may scatter off its surface unpredictably, its trajectory is not greatly affected as it is a massive rigid body. Suppose we study the motion of a small particle, small enough that the random scattering does influence the trajectory and it undergoes Brownian motion: its state will be a probability distribution for position and momentum of the center of mass. Gravitational and inertial forces have not changed, yet the states have changed from ``pure" to statistical ensembles. In other words, the set of states must be closed under both the deterministic evolution of the state and the non-deterministic evolution of the unstated part. If the Brownian motion is not negligible, we do not end in a well defined position/momentum pair, even if we start from one.

@@

As we have hinted, sometimes the state is identified by a distribution (either statistical or actual). Even in this case, the state can be deterministic and reversible. That is, given the distribution at one time we can determine the distribution at future times. The shape and the parameters of the distribution can be deterministic, even if the evolution of the parts are not as they fall within the unstated part. Note that we \emph{cannot} assume trajectories and states are always defined for the unstated part, as this includes also the unknown unknowns. We will return to this aspect when discussing quantum systems.

@@

We now proceed to characterize states and physical distinguishability in more precise terms so that we can capture their description mathematically. What we'll see is that the outcomes of all physical processes that can be used to gain information about the system, that can be used to perform a measurement, induce a topology on the set of states that is Hausdorff and second countable. That is, physical distinguishability is mathematically captured by topological distinguishability. Deterministic and reversible evolution will then preserve the topology, and they will be mathematically captured by self-homeomorphisms.

@@

We can also conceptually combine two different processes into a single one. That is, having a way to measure quantity $x$ and a way to measure quantity $y$ gives us a way to measure the combination $(x,y)$. For ensembles, if we can measure the marginal distribution $\rho_x(x)$ and the marginal distribution $\rho_y(y)$, we know the joint distribution $\rho(x,y)$ has to be compatible with both. Note, though, that this does not provide a way to fully measure $\rho(x,y)$ as we know nothing about the correlation between the two variables.\footnote{The quantum case is similar: we measure marginal distributions and rule out states that are incompatible with them.} Formally, the states compatible with the outcomes of the combined process will be the intersections of the states compatible with each pair of outcomes of the original processes.

@@

\begin{justification}
We claim $\mathcal{S}$ is a set. Each state is well defined as it is physically distinguishable. The collection of all possible states forms a set.

We claim $\mathcal{S}$ has a topology $\mathsf{T}$. Consider the set of all possible physical outcomes associated with all physical processes. Each possible outcome is associated with a set of states that are compatible with that outcome. Let $\mathsf{T}$ be the set of all sets associated with all physical outcomes. $\varnothing \in \mathsf{T}$ and is associated with impossible outcomes. $\mathcal{S} \in \mathsf{T}$ and is associated with unavoidable outcomes. Let $V_1, V_2 \in \mathsf{T}$. Then, by definition, there exists a process $P_1$ that admits $V_1$ as an outcome and a process $P_2$ that admits $V_2$ as an outcome. Consider the process $P$ that combines the outcomes of $P_1$ and $P_2$. This always exists physically as we can prepare the same state multiple times and let it interact with each process separately. $P$ will have a possible outcome $V$ corresponding to the case where $P_1$ gave outcome $V_1$ and $P_2$ gave outcome $V_2$. The states compatible with $V$ must be in both $V_1$ and $V_2$, that is $V = V_1 \cap V_2$. Therefore $\mathsf{T}$ is closed under intersection. Let $V_1, V_2 \in \mathsf{T}$. If they are physically distinguishable, then there exists a physical process $P$ that admits both as outcomes. Given $P$, we can always construct the process $P_0$ that combines $V_1, V_2$ into a single outcome $V$ by "forgetting" which of the two was given. The states compatible with $V$ must be in either $V_1$ or $V_2$, that is $V = V_1 \cup V_2$. Therefore $\mathsf{T}$ is closed under union. $\mathsf{T}$ is a topology by definition.

@@

We claim that $\mathcal{S}$ is second countable. Consider the set of outcomes of a process. As each is identified by a finite number of digits, the set of outcomes is countable. This remains true when combining processes: the number of processes we can combine is finite as is the number of times we can prepare the same state and the set of intersections obtained from the countable outcomes of a finite number of processes is still countable. Consider a basis for the topology. Suppose it is of cardinality greater than countable. Then the basis would be able to distinguish among states in a way that no physical process ever could. The topology could distinguish among states that are not physically distinguishable. The basis for the topology cannot be of cardinality greater than countable. By contradiction, $\mathcal{S}$ must be second countable.

@@

Note that the arguments that led to the topological space had nothing to do with states per se, just that they are physically distinguishable. States, though, are not the only objects with that property. In fact: any element of a set of physical objects (e.g. forces, physical properties such as mass or charge, time) needs to be physically distinguishable to be well defined. We can generalize the above justification: any set of physically distinguishable elements is a topological space.

@@

Note that not all continuous functions correspond to physically meaningful states: we must make sure they describe a finite amount of material.

@@

For each classical composite state will also tell us the total amount of material, once we fixed unit (e.g. mole, kg, ...). That is, there will be a functional $\Lambda : \mathcal{C} \rightarrow \mathbbm{R}$ that tells us how much gas we have for a particular gas configuration. As we combine parts, the total amount of material will be the sum of their parts. As we increase or decrease by a constant factor the amount of material will increase or decrease by the same factor. That is: $\Lambda(a_1 \mathcal{c}_1 + a_2 \mathcal{c}_2) = a_1 \Lambda(\mathcal{c}_1) + a_2 \Lambda(\mathcal{c}_2)$. For a proper state, one that is not the difference of two states, we will expect a positive amount of material. So, if a density $\rho(\mathcal{S})$ is positive everywhere, then the total amount is also positive. This makes $\Lambda$ a positive linear functional.

Being able to tell the total amount is not enough. Experimentally, what we do is measure the amount of material for a region of phase space. For example, the amount of gas within a particular region of space for a particular range of momentum. In general, given an outcome of our physical process (i.e. an open set in the topology for $\mathcal{S}$), each composite state should tell us the amount of material associated with it (i.e. contained in the open set of states). For each $U \subseteq \mathcal{S}$ we'll have a positive linear functional $\Lambda_U : \mathcal{C} \rightarrow \mathbbm{R}$ whose value is always bounded.

@@

In mathematical terms, we can use the Riesz representation theorem for linear functionals to give $\mathcal{S}$ the structure of a measure space, with a suitable Borel $\sigma$-algebra and measure $\mu$.

\begin{defn}\label{integration}
	The state space $\mathcal{S}$ for the particles of a homogeneous classical material is endowed with a measure $\mu$. The state space $\mathcal{C}$ is isomorphic to the space of continuous functions that are integrable over any subset of $\mathcal{S}$. That is: $\mathcal{C} \cong C(\mathcal{S}) \cap  L^1_{ew}(\mathcal{S}, \mu)$ where $L^1_{ew}(\mathcal{S}, \mu) = \{ \rho : \mathcal{S} \rightarrow \mathbbm{R} \; | \; |\int_{U} \rho d\mu| < \infty \; \forall U \subseteq \mathcal{S}\}$.
\end{defn}

@@

Before we can properly identify $d\mu$, it turns out that we need to better characterize $\mathcal{S}$.

On a state space with discrete topology and a single state variable we simply have $\Lambda_U (\mathcal{c}) : \sum \limits_{q=a}^b D(q)$. For one dimensional continuous states, we would simply expect $\Lambda_U (\mathcal{c}) : \int_a^b \rho (q) dq$. By a simple change of state variable, we see that this does not work: $\rho(q')= \rho(q) dq/dq'$. This would make $\rho$ not a function of just the state $\mathcal{s}$, but also of the state variables we are using. This is inconsistent with our previous definition. Moreover if the transformation is not differential, the density is not even well-defined. This tells us that $\mathcal{S}$ cannot be any manifold: it has to be one that allows coordinate independent densities.

To do that $\mathcal{S}$ has to be, first of all, a differentiable manifold. This way, at least, the densities is well defined for all state variables.

If $\rho$ is to remain unchanged when one state variable changes, then one or more other state variables have to change as well so that the Jacobian of the transformation is unitary. This means that we cannot change physical units to all the variables as we please: they are part of a unit system. For example, changing units of position will also change units of momentum. Let $\mathcal{Q}$ be the manifold charted by a the minimum set of state variables that define such system of units. An element of $\mathcal{S}$ will be a geometrical object in such space. As these are states for infinitesimal parts, it will be a local object in $\mathcal{Q}$. As we saw before, it cannot be a point as the density $\rho(q')= \rho(q) dq/dq'$ is not invariant. It cannot be a vector $v(q) e_q$, as the density $\rho(q', v')= \rho(q, v) (dq/dq')^2$ is not invariant. A covector $k(q) dq$ is fine as $\rho(q', k')= \rho(q, k) (dq/dq') (dq'/dq) = \rho(q, k)$. In other words: a covector is the simplest local geometrical object that allows coordinate independent densities. We can identify $\mathcal{S}$ as the cotangent bundle $T^*\mathcal{Q}$, the space of all possible covectors.

If $\mathcal{Q}$ is one dimensional, then the state is fully identified by a pair $(q, k)$. $d\mu$ is proportional to $dq \wedge dk$. We set $d\mu = \hbar dq \wedge dk$, where $\hbar$ is the proportionality constant that will depend on the unit chosen to count the possibilities of the pair $(q, k)$. The cardinality for states in a finite (i.e. compact) region $U \subseteq \mathcal{S}$ will be given by $\int_U \hbar dq \wedge dk$.

If $\mathcal{Q}$ is $n$-dimensional, $d\mu$ is proportional to $dq^n \wedge dk_n$. But it's not just the volume that is preserved as we change state variables. Each pair $(q^i, k_i)$ represents an independent set of state variables, with the cardinality of possibilities given by $\hbar dq^i \wedge dk_i$. We call such a pair a degree of freedom. As they are independent the total number of states is the product of the possibilities of each d.o.f.: $\hbar^n dq^n \wedge dk_n = \prod \limits_{i=1}^n \hbar dq^i \wedge dk_i$. This is equivalent to saying that independent d.o.f. are orthogonal surfaces in $\mathcal{S}$.

The cardinality of possibilities for each d.o.f. (i.e. the wedge product within d.o.f) and the orthogonality of different d.o.f. (i.e. the scalar product across d.o.f.) must be the same regardless of the choice of state variables. We can express both requirements mathematically in a compact way. We first define conjugate momentum as $p_i=\hbar k_i$ and unified state variables as $\xi^a\equiv \{q^i, p_i\}$. Then we consider the canonical two-form $\omega$ given by the following components:
\begin{align*}
\omega_{ab} =  \left[
\begin{array}{cc}
0 & 1 \\
-1 & 0 \\
\end{array}
\right] \otimes I_n =
\left[
\begin{array}{cc}
0 & I_n \\
-I_n & 0 \\
\end{array}
\right] \\
\end{align*}
It returns the wedge product within a d.o.f. and the scalar product across. Requiring the invariance of this metric under state variables assures us the cardinalities of states and possibilities is well defined.

The state space $\mathcal{S}$ is the symplectic manifold $(T^*\mathcal{Q}, \omega)$. This is the simplest manifold that allows us to define coordinate invariant densities. The metric allows us to measure the cardinality of possibilities within and across independent d.o.f.

\begin{defn}\label{classical_phase_space}
The state space $\mathcal{S}$ for the particles of a homogeneous classical  system is a symplectic manifold form by a cotangent bundle $T^*\mathcal{Q}$ equipped with the canonical two-form $\omega$.
\end{defn}

\begin{justification}
	TODO: we can search for omega looking from constraints. Must be linear. Must be skew-symmetric (applied to the same vector twice must be zero). Must be invariant by coord transformation.
\end{justification}

We should note that we are starting to be in a position similar to the one discussed in \ref{discrete_state_metric} for discrete states. We have both a topological space and a metric that allows to count states and possibilities. The only thing we started from was the idea of an infinitesimally reducible system. That lead to a state space $\mathcal{S}$ for the infinitesimal parts, on which we must be able to define state variable independent distributions. Which in turn gave us degrees of freedom made by pair of conjugate variables and the symplectic metric typical of classical phase space.

Before discussing time evolution, we should investigate what assuming our state is infinitesimally reducible means in terms of the fundamental model discussed in Assumption \ref{detrevass}.
